\documentclass[11pt]{article}

\usepackage{amsmath, amssymb, geometry}
\geometry{margin=1in}

\title{Dynamic Low-Rank Fencing Rating Model}
\date{}

\begin{document}
\maketitle

\section{Observation Model}

For a bout between fencers $i$ and $j$ during month $m$, define the normalized score differential
\[
Z_{ijm}
=
\frac{\text{score}_i-\text{score}_j}{\text{target score}},
\]
where the target score is $5$ for pool bouts and $15$ for direct-elimination bouts.

We model
\[
Z_{ijm}
=
\mu_{ijm}+\varepsilon_{ijm},
\]
with
\[
\varepsilon_{ijm}\sim \mathcal N(0,\sigma^2_{\mathrm{type}}),
\]
where the noise variance may differ between pool and DE bouts.

\section{Latent Performance Model}

The conditional mean is
\[
\boxed{
\mu_{ijm}
=
s_{im}-s_{jm}
+
u_i^\top v_j-u_j^\top v_i
+
\beta_{\mathrm{age}}(a_{im}-a_{jm})
+
c_{\ell(i)}-c_{\ell(j)}
+
\beta_{\mathrm{DE}}\mathbf 1\{\mathrm{DE}\}.
}
\]

Here:
\begin{itemize}
    \item $s_{im}$ is the latent skill of fencer $i$ in month $m$.
    \item $u_i,v_i\in\mathbb R^k$ are stable low-dimensional matchup factors.
    \item $u_i^\top v_j-u_j^\top v_i$ is an antisymmetric low-rank matchup effect.
    \item $a_{im}$ is the age of fencer $i$ in month $m$.
    \item $\ell(i)$ is the club of fencer $i$.
    \item $c_{\ell}$ is the club effect for club $\ell$.
    \item $\mathbf 1\{\mathrm{DE}\}$ indicates a direct-elimination bout.
\end{itemize}

\section{Dynamics}

Only skill is allowed to vary month to month. The matchup factors are stable.

Smoothness is imposed through
\[
\lambda_{\mathrm{time}}
\sum_i\sum_m
\left(s_{im}-s_{i,m-1}\right)^2.
\]

Thus the model allows rapid but not arbitrary month-to-month skill movement.

\section{Estimation}

Estimate parameters by minimizing
\[
\sum_b
\left(Z_b-\mu_b\right)^2
+
\lambda_{\mathrm{time}}
\sum_i\sum_m
\left(s_{im}-s_{i,m-1}\right)^2
+
\lambda_s\sum_{i,m}s_{im}^2
\]
\[
+
\lambda_u\sum_i \|u_i\|^2
+
\lambda_v\sum_i \|v_i\|^2
+
\lambda_c\sum_\ell c_\ell^2
+
\lambda_\beta \|\beta\|^2.
\]

Equivalently, if using bout-type-specific variances, minimize
\[
\sum_b
\frac{
\left(Z_b-\mu_b\right)^2
}{
2\sigma_{\mathrm{type}(b)}^2
}
+
\text{regularization terms}.
\]

\section{Identifiability}

Because skill is relative, impose
\[
\sum_i s_{im}=0
\qquad \forall m.
\]

Because club effects are also relative, impose
\[
\sum_\ell c_\ell=0.
\]

The low-rank factors $u_i,v_i$ are not individually identifiable up to rotations and transformations, but the induced matchup residual
\[
r_{ij}=u_i^\top v_j-u_j^\top v_i
\]
is the object of interest.

\section{Interpretation}

The model decomposes observed performance into three parts:
\[
\text{score differential}
=
\text{dynamic skill difference}
+
\text{stable matchup effect}
+
\text{covariate effects}
+
\text{noise}.
\]

The central modeling choice is that
\[
s_{im}
\]
evolves by month, while the non-transitive matchup structure
\[
u_i^\top v_j-u_j^\top v_i
\]
is stable over time.

This keeps the model flexible enough to capture stylistic non-transitivities, but not so flexible that each fencer's style must be relearned every month.

\end{document}
